Symbolic Abstract Data Type Inference

### 2015-07-09 algorithm.tex
Algorithm cleanup.

Algorithm~\ref{alg:inference} lists the algorithm for solving the symbolic ADT
inference problem which is based on the relation $\preceq$ defined in
Section~\ref{sec:patterns}. The algorithm performs an weight-increasing
enumeration of $B$ while maintaining the $\preceq$-minimal elements, and stops
whenever no new minimals are found at a given weight.

\begin{algorithm}[t]
  \SetAlgoLined
  \SetKwInOut{Input}{Input}
  \Input{A reference implementation $\mathcal{I}$}
  \KwResult{A formula representing the ADT of $\mathcal{I}$}
  weight $\gets$ $0$ \;
  patterns $\gets$ $\emptyset$ \;
  \Repeat{$B_\text{weight} = B_{\text{weight}-1}$}{
    weight $\gets$ weight + $1$ \;
    patterns $\gets$ $\min B_\text{weight}$ \;
  }
  \Return $F(\text{patterns})$
  \caption{Symbolic ADT inference.}
  \label{alg:inference}
\end{algorithm}

The partial correctness of Algorithm~\ref{alg:inference} (i.e.,~ignoring
termination) relies on the input ADT being \emph{predictable}, i.e.,~for all
$i \in \mathbb{N}$,
\begin{align*}
  B_{i+1} \succeq \min \bigcup_{0 < j \le i} b_j
  \text{ implies } \min B = \min \bigcup_{0 < j \le i} b_j
\end{align*}
Section~\ref{sec:nature} shows that many naturally-occurring ADTs are predictable.

@@

\begin{proof}
  Termination is a direct consequence of Lemma~\ref{lem:wqo}, and since the
  number of histories of any weight $i \in \mathbb{N}$ is finite\footnote{
    As noted in Section~\ref{sec:histories}, we consider equality between
    histories up to renaming of operation identifiers.
  }
  each $B_i$ is computable.

  Furthermore, when $A$ is predictable, we
  have $B \subseteq \text{patterns}^+$, where $\text{patterns}^+$ is the
  upward closure of $\text{patterns}$ under $\preceq$. Thus by
  Lemma~\ref{lem:formula1}, the returned formula represents $A$.
\end{proof}

Note that for non-predictable ADTs, the value of $\text{patterns}^+$ upon
termination of Algorithm~\ref{alg:inference} is an under-approximation of the
bounded ADT complement $B$. The resulting symbolic ADT representation is still
satisfied by all histories in $A$, therefore it would still be sound for modular
program reasoning, identifying only actual violations, and implying the
correctness of client programs which do not depend on the stronger criteria
which excludes unidentified violations. However, the under-approximation may be
incomplete in identifying all violations, and in proving client programs which
depend on the stronger criteria which excludes them all.

>>>>

Algorithm~\ref{alg:inference} solves the symbolic ADT inference problem by
computing a finite representation of an ADT complement $B$ using the $\preceq$
relation of Section~\ref{sec:patterns}. In general, this is achieved by
enumerating the elements of $B$ while maintaining the $\preceq$-minimal
elements, and recognizing a condition under which all the elements of $B$ are
related to the current set of minimals~\cite{conf/lics/AbdullaCJT96,
journals/tcs/FinkelS01}. In our case, we stratify our enumeration of $B$ by the
relative sizes of its histories. Formally, we say the \emph{weight} of a history
is the maximum among operation frequencies and the number of matches. We then
define
\begin{align*}
  & B_i = \set{ h \in B : \weight(h) \le i } \\
  & B_i' = \set{ h \in B : \exists h' \in B_i.\ h' \preceq h \text{ and } \weight(h) \le i+1 }
\end{align*}
respectively as the histories of $B$ with at most $i$ matches and duplicates,
and those derived from $B_i$ with at most $i+1$ matches and duplicates. We say
that an ADT with complement $B$ is \emph{predictable} if $B_i^* = B$ whenever
$B_i' = B_{i+1}$, i.e.,~if all histories of $B$ are represented by $B_i$ whenever
all histories of $B_{i+1}$ are represented by $B_i$. Algorithm~\ref{alg:inference}
then performs a weight-increasing enumeration of $B$, collecting
$\preceq$-minimals from smaller-weight violations before advancing to greater
weights. When no violation is found at a given weight, the algorithm terminates.
This algorithm is guaranteed to terminate since $\preceq$ is a
wqo~\cite{conf/lics/AbdullaCJT96, journals/tcs/FinkelS01}. Furthermore, this
algorithm is sound for arbitrary ADTs, and complete for predictable ADTs.
Many naturally-occurring ADTs are predictable — in fact all of the examples we
know of are predictable, as demonstrated in Section~\ref{sec:nature}.

\begin{algorithm}[t]
  \SetAlgoLined
  \SetKwInOut{Input}{Input}
  \Input{A reference implementation $\mathcal{I}$ of width $k$}
  \KwResult{A formula representing the ADT of $\mathcal{I}$}
  patterns $\gets$ $\emptyset$ \;
  $w$ $\gets$ $0$ \;
  \Repeat{none-found}{
    none-found $\gets$ true \;
    \For{each $k$-width history $h$ with weight $w$}{
      \uIf{$h$ is \emph{executable} with $\mathcal{I}$}{
        continue
      }
      \uElseIf{$\exists h' \in \text{patterns}.\ h' \preceq h$}{
        continue
      }
      \Else{
        add $h$ to patterns \;
        none-found $\gets$ false
      }
    }
    $w$ $\gets$ $w$ + $1$ \;
  }
  \Return $F(\text{patterns})$
  \caption{Symbolic ADT inference.}
  \label{alg:inference}
\end{algorithm}

@@

\begin{proof}
  Termination is a direct consequence of Lemma~\ref{lem:wqo}, and since the
  number of histories of any weight $i \in \mathbb{N}$ is finite\footnote{
    As noted in Section~\ref{sec:histories}, we consider equality between
    histories up to renaming of operation identifiers.
  }
  each $B_i$ is computable.
  When $A$ is predictable, then $B = \text{patterns}^*$, and thus by
  Lemma~\ref{lem:formula1}, the returned formula $F(\text{patterns})$
  represents $A$.
\end{proof}

For non-predictable ADTs, the value of $\text{patterns}^*$ upon termination of
Algorithm~\ref{alg:inference} is an under-approximation of the bounded ADT
complement $B$. The resulting symbolic ADT representation is still satisfied by
all histories in $A$, therefore it would still be sound for modular program
reasoning, identifying only actual violations, and implying the correctness of
client programs which do not depend on the stronger criteria which excludes
unidentified violations. However, the under-approximation may be incomplete in
identifying all violations, and in proving client programs which depend on the
stronger criteria which excludes them all.

### 2015-07-09 formula.tex
Algorithm cleanup.

We obtain a formula representing an ADT $A$ by taking the conjunction of the
negative embedding formulas $\set{ \lnot\textsc{Embed}_h : h \in H }$ from any
set $H$ that \emph{generates} the bounded complement $B$ of $\ker A$, i.e.,
$H\subseteq B\subseteq Cl_\preceq(H)$ where $Cl_\preceq(H)$ is the closure of $H$ w.r.t. $\preceq$.

\begin{lemma}\label{lem:formula2}

  The formula $F(H)=\bigwedge_{h \in H} \neg \textsc{Embed}_h$ represents $A$,
  for any set $H$ which generates the bounded complement of $\ker A$.

\end{lemma}
\begin{proof}
For a set $\Sigma$, let $\neg \Sigma$ denote its complement.
By Lemma~\ref{lem:ADT_closure}, the complement of $A$ is closed under $\preceq$. Since
$H\subseteq B\subseteq  \neg \ker A \subseteq \neg A$, we have that $Cl(H)\subseteq \neg A$.
By Lemma~\ref{lem:formula1}, the formula $F(H)$ holds exactly for the set of histories $\neg Cl(H)$.
%which is the complement of $Cl(M)$.
By the previous inclusions, we have that $A\subseteq \neg Cl(H)\subseteq \neg B$.
Therefore, $F(H)$ holds for all histories in $\ker A$
and doesn't hold for histories in $B$, which concludes the proof.
\end{proof}

>>>>

We obtain a formula representing an ADT by taking the conjunction of the
negative embedding formulas $\set{ \lnot\textsc{Embed}_h : h \in H }$ from any
set $H$ that generates its bounded complement $B$, i.e.,~whose closure under
$\preceq$, denoted by $H^*$ henceforth, is equal to $B$. We define the
\emph{exclusion formula} of a set of histories $H$ as $F(H) = \bigwedge_{h \in
H} \neg \textsc{Embed}_h$.

\begin{lemma}
  \label{lem:formula2}

  Let $B$ be the bounded complement of a bounded-width ADT $A$, and $H$
  a set of histories. $F(H)$ represents $A$ if $H^* = B$.

\begin{proof}

  For a set $\Sigma$, let $\neg \Sigma$ denote its complement. By
  Lemma~\ref{lem:ADT_closure}, the complement of $A$ is closed under $\preceq$.
  Since $H\subseteq B\subseteq  \neg A \subseteq \neg A$, we have that
  $H^* \subseteq \neg A$. By Lemma~\ref{lem:formula1}, the formula $F(H)$ holds
  exactly for the set of histories $\neg H^*$. By the previous inclusions, we
  have that $A\subseteq \neg H^* \subseteq \neg B$. Therefore, $F(H)$ holds for
  all histories in $A$ and doesn't hold for histories in $B$.
\end{proof}

### 2015-07-09 overview.tex
Algorithm cleanup.

The key technical obstacle in realizing the abstract algorithm of
Algorithm~\ref{alg:abstract}, overcome in Section~\ref{sec:patterns}, is the
classification of violations into a finite set of patterns, rendering the
remaining invalid sequences redundant. The symbolic representation which
excludes a given set of patterns is relatively straightforward to construct, and
is given in Section~\ref{sec:formula}. Note however that the termination of this
abstract algorithm is nondeterministic, and thus the inferred ADT specification
is generally not the strongest possible in the sense that it may not exclude
certain invalid linearizations. Nevertheless, specifications inferred by this
algorithm are sound, in the sense that when modular program reasoning with
inferred specifications succeeds, correctness follows. In any case, the
specifications inferred by our refinement of this abstract algorithm in
Section~\ref{sec:algorithm} are both sound and complete for the
naturally-occurring ADTs of interest. The remainder of this article develops
the technical machinery required to realize this abstract inference algorithm.

>>>>

The key technical obstacle in realizing the abstract algorithm of
Algorithm~\ref{alg:abstract}, overcome in Section~\ref{sec:patterns}, is the
classification of violations into a finite set of patterns, rendering the
remaining invalid sequences redundant. The symbolic representation which
excludes a given set of patterns is relatively straightforward to construct, and
is given in Section~\ref{sec:formula}. Note however that the termination of this
abstract algorithm is nondeterministic, and thus the inferred ADT specification
is generally not the strongest possible in the sense that it may not exclude
certain invalid linearizations. Nevertheless, specifications inferred by this
algorithm are sound, in the sense that when modular program reasoning with
inferred specifications succeeds, correctness follows. In any case, in
Section~\ref{sec:algorithm} we give conditions under which our refinement
of this abstract algorithm is both sound and complete, and show that these
conditions hold for naturally-occurring ADTs, thus resulting in sound and
complete symbolic ADT specifications. The remainder of this article develops
the technical machinery required to realize this abstract inference algorithm.
